Efficient pedigree recording for fast population genetics simulation

### 2017-12-22 forwards_paper.tex
tiny tweak
add in benchmark results to manuscript

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

@@

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random.
% (and all such instances is independent).  % this is understood here
Given a node table $\Nt$, the function $\taddnode{\Nt, t}$
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node.
Similarly, the function $\taddedge{\Et, \ell, r, p, c}$
adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described in section \ref{ss:simplify})
simplifies the history stored in the tables $\Nt$ and $\Et$
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
A step-by-step explanation follows the pseudocode.

@@

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
neutral mutations (instead of the ARG), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  ARG tracking
dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
the ARG, the relative speedup due to ARG tracking is up to $\approx 50$ fold (Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, ARG tracking reduces runtimes more for larger population sizes
(Figure~\ref{fig:relative_speedup_selection}).  We could not make such an observation with \simupop{}, which tended to
run out of memory on our system.  Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
partially obscures an important observation that the run times appear approximately linear in region size when ARG
tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on average
each generation, and the number of edges is a function of $\rho$ in expectation.

We also performed a more limited set of simulations without natural selection.  The total run times are shown in
Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to ARG tracking is again substantial in these
simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to ARG tracking here with \fwdpy11{}
than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
and~\ref{sfig:speedup_nosel}.

\begin{figure}
    \includegraphics[]{sims/rawspeed}
    \caption{\label{fig:runtimes_selection}Total run time for a single simulation replicate as a function of region
        length, measured as the scaled recombination parameter $\rho = 4Nr$. The different colored lines represent
        different diploid population sizes ($N$). The left column shows run times for
        simulations including the tracking of neutral mutations.  Missing data points are due to either simulations
    being killed at 72 hours or running out of memory (which only happened with \simupop{}).  The right column of
panels shows run times tracking the ARG instead of the neutral mutations. The dashed line in the upper right panel shows
the results for an implementation using \fwdpy{} where the ARG simplification steps were handled in a separate thread of
execution and fitness calculations were parallelized.}
\end{figure}

\begin{figure}
    \includegraphics[]{sims/speedup}
    \caption{\label{fig:relative_speedup_selection}Relative performance improvement due to ARG tracking. This figure
    shows the ratio of the run time tracking neutral mutations over the run time tracking the ARG. Data points are taken
from Figure~\ref{fig:runtimes_selection} for simulations that ran to completion using both methods.}
\end{figure}

@@

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and all such instances are independent).
Given a node table $\Nt$, the function $\taddnode{\Nt, t}$
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node.
Similarly, the function $\taddedge{\Et, \ell, r, p, c}$
adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described in section \ref{ss:simplify})
simplifies the history stored in the tables $\Nt$ and $\Et$
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
A step-by-step explanation follows the pseudocode.

@@

\newpage
\clearpage
\section*{Supplementary Information}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[!h]
    \includegraphics[]{sims/rawspeed_nosel}
    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
    selection.}
\end{figure}

\begin{figure}
    \includegraphics[]{sims/speedup_nosel}
    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to ARG tracking for simulations without
    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
neutral variation over the run times tracking the ARG.}
\end{figure}
